Notes and other text

### 2023-03-31 doc/manuscript/translocation.qmd
Expand on key points in the discussion

Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
"Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014] <!--# Bree comment: Nuances of word choice -->.
In host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021]<!--# The previous sentence might fit better in the Discussion when we take a deeper dive into evolutionary rescue -->.
Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, is difficult and remains rare.
In addition to promoting recovery of remnant populations after disease-induced declines, selection that promotes recovery of focal populations can also provide sources of individuals with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
To date, the use of resistant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
<!--# Other examples in addition to Joseph and Knapp (2018)? In other taxa besides amphibians? -->

@@

In the current study, we combine results from novel <!--# better adjective than "novel" --> genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed reduced susceptibility/increased resistance to Bd infection observed in recovering populations [@vredenburg2010] <!--# How best to make the jump from "reduced susceptibility" to "increased resistance"? -->.
We then reintroduce these resistant frogs to 24 sites and describe the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant animals can allow the large-scale recovery of disease-impacted species, with potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.

@@

1.  Summary of all of the results in a way that integrates the different components/sections (selection, translocations, viability), and makes the pitch for how unique the results are (as per comment from Bree):
    -   Evidence for evolutionary rescue in a conservation-relevant species.
    -   Successful reestablishment of extirpated populations via translocation despite ongoing Bd infection.
    -   Evidence for long-term viability of reestablished populations.

<!-- -->

2.  Evolutionary rescue: We provide five lines of evidence that evolutionary rescue is a strong candidate mechanism for observed resistance.
    -   Epidemiological conditions in MYL-Bd systems X, Y, and Z are conducive to evolutionary rescue.
    -   Differences in allele frequencies exist between adjacent Bd-naive and recovering populations.
    -   There is a shared selective response across recovering populations.
    -   Changes in population growth rate due to evolution extend beyond the natal environment.
    -   In populations established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).

>>>>

Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
"Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014] <!--# Bree comment: Nuances of word choice -->.
Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, is difficult and remains rare.
In addition to promoting recovery of remnant populations after disease-induced declines, selection that promotes recovery of focal populations can also provide sources of individuals with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
To date, the use of resistant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
<!--# Other examples in addition to Joseph and Knapp (2018)? In other taxa besides amphibians? -->

@@

In the current study, we combine results from novel <!--# better adjective than "novel". Mark here: Reads OK without an adjective --> genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed reduced susceptibility/increased resistance to Bd infection observed in recovering populations [@vredenburg2010] <!--# How best to make the jump from "reduced susceptibility" to "increased resistance"? Mark here:  This transition reads fine to me. -->.
We then reintroduce these resistant frogs to 24 sites and describe the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
We use these data to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant animals can allow the large-scale recovery of disease-impacted species, with potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.

@@

Disease-induced declines are a central conservation challenge for wildlife and identifying biologically grounded strategies for assisted recovery is a critical need across systems.
One such strategy requires leveraging disease-induced genetic changes in host populations and so-called evolutionary rescue to recover a species at the landscape scale.
While this strategy has been previously proposed, it has never been successfully implemented.
Here, we provide strong evidence that selection for resistance and evolutionary rescue has led to the recovery of some populations of the endangered MYL frog in the presence of a deadly fungal pathogen.
Leveraging individuals from these resistant and rescued populations, we use 15 years of translocation effort and population viability modeling to demonstrate that populations that have experienced evolutionary rescue can be used to successfully establish and recover this species across the landscape.
These results not only provide a rare example of species recovery following disease-induced declines due heritable changes in resistance, but provide the first demonstration of how natural evolutionary rescue can be leveraged to successfully recover a species in the presence of disease.

While conclusively demonstrating that evolutionary rescue is driving population recoveries is challenging.
we provide four lines of evidence that evolutionary rescue is a strong candidate mechanism for observed resistance and recoveries in MYL frog systems.
First, in host-pathogen systems, evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
At least two of these conditions are met in the MYL frog system: i) individuals that recover from Bd infection do not gain long-lasting immunity to Bd and are susceptible to re-infection and ii) there are potential Bd reservoirs (Briggs, Reeder) and seasonal frequency-dependent transmission dynamics (Wilber) in the MYL frog system.
Thus, MYL frog-Bd systems have the epidemiological characteristics necessary for evolutionary rescue to occur.
Second, our genomic analyses comparing naive and recovering MYL frog populations identified notable differences in allele frequencies between adjacent Bd-naive and recovering population.
These differences are consistent with strong signatures of selection and regions of the genome under selection are associated with known mechanisms of resistance to Bd infection.
Third, our translocations allowed us to conclude that observed resistance and recovery of resistant populations is not strictly due to local environmental conditions (though environment does play a role).
Changes in population growth rate due to evolved resistance extend beyond the natal environment and into subsequent generations, as demonstrated by high survival rates of translocated frog populations in new habitats in the presence of Bd.
Finally, evolved resistance has a quantifiable phenotype.
In persistent populations and those established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics).
While other studies have documented signatures of selection in hosts following disease-induced declines (Epstein, white-nose, Byrne), our study is unique in that we show that evolved resistance is associated with population recovery in both natal and non-natal habitats, providing an actionable strategy for promoting the long-term viability of a species.

Despite the results we present here, recovery of Bd-impacted amphibians worldwide is expected to be challenging at best.
Hundreds of declining species, but mostly failed attempts at recovery.
Failures may be due in part to a general inability by amphibians to mount an effective immune response against Bd infection.
In contrast, in our study MYL frogs (a species highly susceptible to chytridiomycosis) showed relatively high frog survival despite ongoing Bd infection and low probability of extinction for many reestablished populations.
